# Regression Analysis
## Tasks

*Describe the work you have done this week and summarize your learning.*
- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

*Preamble*
```{r}
# load packages
library(dplyr)
library(ggplot2)
library(GGally)
`````

**Q1: Loading Dataset**
```{r}
learning2014 <- read.csv("~/R/win-library/4.0/IODS-project/learning2014.csv")
str(learning2014)
dim(learning2014)
date()
```
The data set currently contains 166 observations (information about 166 students) and 7 variables (age, gender, attitude, deep, stra, points). This came from a larger data file containing 183 observations with 60 variables. Please refer to [this link](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-meta.txt) to find more details.

**Q2: Visualization and Summary**
```{r}
# visualization of the data set
# code from datacamp
ggpairs(learning2014, mapping=aes(col=gender, alpha=0.3), lower=list(combo=wrap("facethist", bins=20)))
# summary the data set
summary(learning2014)
```

From data visualizations there are no strong correlations between the variables of interest (columns). The highest positive correlation is between attitude and points (44%) whiles the highest negative correlation is between deep and surf (-32%). The distributions are similar to Gaussian distributions. We see more females in the data than males.


