# Regression Analysis
## Tasks

*Describe the work you have done this week and summarize your learning.*
- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

*Preamble*
```{r}
# load packages
library(dplyr)
library(ggplot2)
library(GGally)
`````

**Q1: Loading Dataset**
```{r}
learning2014 <- read.csv("~/R/win-library/4.0/IODS-project/learning2014.csv")
str(learning2014)
dim(learning2014)
date()
```
The data set currently contains 166 observations (information about 166 students) and 7 variables (age, gender, attitude, deep, stra, points). This came from a larger data file containing 183 observations with 60 variables. Please refer to [this link](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-meta.txt) to find more details.

**Q2: Visualization and Summary**
```{r}
# visualization of the data set
# code from datacamp
ggpairs(learning2014, mapping=aes(col=gender, alpha=0.3), lower=list(combo=wrap("facethist", bins=20)))
# summary the data set
summary(learning2014)
```

From data visualizations there are no strong correlations between the variables of interest (columns). The highest positive correlation is between attitude and points (44%) whiles the highest negative correlation is between deep and surf (-32%). The distributions are similar to Gaussian distributions. We see more females in the data than males.

**Q3 & Q4: Fit A Regression Model**
```{r}
# fit a multiple linear regression model
reg_model = lm(points ~ attitude + stra + surf, data=learning2014)

# summary the fitted model
summary(reg_model)
```
According to the summary of the fitted model, we could see that points=11.0171+3.3952∗attitude+0.8531∗stra−0.5861∗surf. stra and surf do not show a statistically significant relationships with points.

F-statistic (p-value) is used to check the significance.

```{r}
# remove stra and surf to fit a new model
reg_model1 = lm(points ~ attitude, data=learning2014)
# summary the new model
summary(reg_model1)
```

According to the summary of the new fitted model, we could see that points=11.6372+3.5255∗attitude. The multiple R squared measures how much of the variability in the dependent variable (points) is caused by the variability in the independent variables (attitude, stra and surf). In model 1 it is 21% and in model 2 it is 19.1%. In the first model, there is a prediction that all other things held constant as points increase by 1 score, attitude and stra increase by 3.4 and 0.9 respectively but surf reduces by 0.6. In the second model, there is a prediction that as points increase by 1 score, attitude will increase by 3.5, holding other factors constant.