# Logistic regression
## Tasks

*Data Wrangling and Analysis.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

*Preamble*
```{r}
# load packages
library(dplyr)
library(ggplot2)
library(GGally)
library(openxlsx)
`````

**Q2: Loading Dataset**
```{r}
pormath <- read.xlsx("~/R/win-library/4.0/IODS-project/data/pormath.xlsx")

colnames(pormath)
date()
```
This data approach student achievement in secondary education of two Portuguese schools. The data attributes include student grades, demographic, social and school related features) and it was collected by using school reports and questionnaires. Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). Please refer to [this link](https://archive.ics.uci.edu/ml/datasets/Student+Performance) to find more details.
