# Clustering and Classification
## Tasks

*Analysis.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

*Preamble*
```{r}
# load packages
library(dplyr)
library(ggplot2)
library(GGally)
library(tidyr)
library(corrplot)
`````

**Q2: Loading Dataset**
```{r}
# access the MASS package
library(MASS)

# load the data
data("Boston")

# explore the dataset
str(Boston)
dim (Boston)
```
This dataset describes the housing values in the Suburbs of Boston. It contains 506 observations with 14 variables. Please refer to [this link](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html) to find more details about the variables.

**Q3:Data overview**
```{r}
#Summaries
summary(Boston)

# calculate the correlation matrix and round it
cor_matrix<-cor(Boston)
cor_matrix<-cor(Boston)%>% round(digits = 2)

# print the correlation matrix
cor_matrix

# visualize the correlation matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)

```

From the correlation values and plots we can tell the relationships between the variables. There is high positive correlation between  the index of accessibility to radial highways (rad) and full-value property-tax rate per \$10,000 (tax) but negative correlation between lower status of the population (lstat) and median value of owner-occupied homes in \$1000s (medv).

**Q3:Standardize data**
```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled<-as.data.frame(boston_scaled)

# summary of the scaled crime rate
summary(boston_scaled$crim)

# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)
```
